---
title: "Assignment 3: Data Exploration"
author: "Rachael Stephan"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Setup,  message = FALSE}
#load in packages
library(tidyverse); library(lubridate); library(here)

#check working directory
getwd()

#check current location
here()

#upload data sets with strings as factors
Neonics <- read.csv(
  file = here('./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = TRUE)

Litter <- read.csv(
    file = here('./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
    stringsAsFactors = TRUE)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonictinoids are a common pesticide that have recently been suspect of having deleterious effects on insect populations, including acting as neurotoxins. Neonictinoids have a high toxicity for invertebrates, and their broad spectrum application mean they will impact many taxa. This can become an issue when the pesticides begin to have impacts on nontarget species that provide important ecosystem services, such as pollinators. Understanding how common agriculture pesticides impact invertebrates can help us preserve healthy populations.  

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: litter and woody debris may be important components of the carbon cycle and ecosystem component. The litter and woody debris will release carbon and organic matter into the soil as they decompose. Certain animals may also use them as a food source (i.e., decomposers) or as a habitat. Additionally, the litter and woody debris may pose as a fire risk as they can act as the tinder to flames.   

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litterfall and woody debris are quantified by mass captured in aerial and ground traps 
 2. Masses < 0.01 g are reported for presence of functional groups but not quantified
 3. finest temporal resolution is day of trapping; finest spatial resolution is one trap
 4. latitude, longitude, and elevation of the plot center are recorded

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Neonics_dim}
#get dimensions of Neonics
dim(Neonics)

```
> Answer: 
4623 rows and 30 columns

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Neonics_sum_effect}
#get summary statistics of the effects of neonictinoids and sort them from high to low
sort(summary(Neonics$Effect),decreasing = TRUE)
```

> Answer: The top 5 most common effects from most common to less common are population, mortality, behaviour, feeding behaviour, and reproduction. These effects may be of specific interest because it is important to understand the main effects of the pesticides. Also, looking at commanalities between these effects can indicate severity and possible target systems of the insecticide. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Neonics_sum_species}
#get summary statistics of the studied species
summary(Neonics$Species.Common.Name, maxsum = 7) #note: sort is not needed with maxsum because maxsum automatically takes the most frequent levels. n+1 levels need to be specified because one is dedicated to "other".

```

> Answer: The top six species are all bees or wasps. These are all important pollinators. Pollinators perform important ecosystem services of pollinating plants to ensure their successful reproduction. Without these species, or with impaired individuals, plant reproduction would falter and the parts of the ecosystem they support would also suffer.  

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Neonics_conc}
#determine the class of the `Conc.1..Author.` column
class(Neonics$Conc.1..Author.)

```

> Answer: The column has a class of factor. The column represents the concentration of the chemicals, but the units of these concentrations are not all the same. Therefore, it would not be helpful to have all of these values on the same numerical axis because their relative relationships would not be accurate. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Neonics_polygeom}
#create a frequency polygon graph to show how many studies were conducted per year
ggplot(Neonics)+
  geom_freqpoly(aes(x = Publication.Year))+
  labs(
    title = "Number of Publications on the Effect of Neonictinoids on\nInsects by Year",
    x = "Publication Year",
    y = "Count")

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Neonics_polygeom_colour}
#create a frequency polygon graph as above but specify the colour is based on location
ggplot(Neonics)+
  geom_freqpoly(aes(x = Publication.Year, colour = Test.Location))+
  labs(
    title = "Number of Publications on the Effect of Neonictinoids on\nInsects by Year",
    x = "Publication Year",
    y = "Count")

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test location is the laboratory. Although laboratory studies have always been popular, there were some periods of time (i.e., ~1993-2001 and 2009) where natural field studies were more popular (or published). Artifical field and undeterminable field studies have not been very popular over this data set. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Neonics_bar}
#create a bar graph that shows the common endpoints with the most common at the origin
ggplot(Neonics)+
  geom_bar(aes(x = fct_infreq(Endpoint)))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(
    title = "Counts of Endpoints in Neonictinoid Studies on Insects",
    x = "Endpoint",
    y = "Count")
```

> Answer: FILL IN ANSWER HERE. ANSWER. ANSWER. ANSWER





















## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}


```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}


```

> Answer:

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
